---
title: "Real Income vs Housing Prices in Singapore"
subtitle: "Did income growth keep pace with HDB resale prices, 2000–2025?"
author:
  - name: Shina Shih Xin Rong (2402781)
  - name: Jeanie Cherie Chua Yue-Ning (2403083)
  - name: Claudia Yue Xin Ying (2401817)
  - name: Nurul Zahirah Binte Muhamadnoh (2401671)
  - name: Anushka Chourasia (2503769)
date: "2026-03-22"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
    number-sections: false
    theme: cosmo
    css: styles.css
execute:
  echo: false
  warning: false
  message: false
---
```{r setup}
library(tidyverse)
library(knitr)
library(scales)

tidy_data <- readRDS("tidy_data.rds")

gap_data <- tidy_data %>%
  pivot_wider(names_from = Measure_Type, values_from = Index_Value) %>%
  mutate(Affordability_Gap = `HDB Resale Price` - `Real Income`)

col_measure <- c("Real Income" = "#2166AC", "HDB Resale Price" = "#B2182B")
col_group   <- c("Bottom 20%" = "#D73027", "Middle 60%" = "#FC8D59", "Top 20%" = "#4575B4")

base_theme <- theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 14, margin = margin(b = 6)),
    plot.subtitle    = element_text(size = 10, color = "grey40", margin = margin(b = 12)),
    plot.caption     = element_text(size = 8, color = "grey55", hjust = 0,
                                    margin = margin(t = 10)),
    plot.background  = element_rect(fill = "#FAFAFA", color = NA),
    panel.background = element_rect(fill = "#FAFAFA", color = NA),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey90"),
    legend.position  = "bottom",
    legend.key.width = unit(1.5, "cm"),
    strip.text       = element_text(face = "bold", size = 12, margin = margin(b = 6)),
    axis.text        = element_text(color = "grey30"),
    axis.text.x      = element_text(angle = 45, hjust = 1, size = 9)
  )
```

## The Improved Visualization
```{r main-chart, fig.width=13, fig.height=6}
labels_2025 <- tidy_data %>%
  filter(Year == 2025) %>%
  mutate(label = paste0(round(Index_Value, 0)))

events <- tibble(
  xmin  = c(2008, 2020),
  xmax  = c(2009, 2021),
  label = c("GFC", "COVID-19")
)

p <- ggplot(tidy_data,
            aes(x = Year, y = Index_Value, color = Measure_Type,
                linetype = Measure_Type)) +

  # Shaded gap ribbon
  geom_ribbon(
    data = gap_data,
    aes(x = Year, ymin = `Real Income`, ymax = `HDB Resale Price`),
    inherit.aes = FALSE,
    fill = "#B2182B", alpha = 0.07
  ) +

  # Event shading
  geom_rect(data = events,
            aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
            inherit.aes = FALSE,
            fill = "grey50", alpha = 0.10) +

  # Event labels — inside grey band, rotated, at bottom
  geom_text(data = events,
          aes(x = (xmin + xmax) / 2, y = 58, label = label),
          inherit.aes = FALSE,
          size = 2.5, color = "grey35", fontface = "italic",
          angle = 90) +

  # Base year reference line
  geom_hline(yintercept = 100, linetype = "dashed",
             color = "grey55", linewidth = 0.45) +

  # Main lines
  geom_line(linewidth = 1.1) +
  geom_point(data = tidy_data %>% filter(Year %% 5 == 0),
             size = 2.2, shape = 21, fill = "white", stroke = 1.1) +

  # 2025 endpoint labels
  geom_text(
    data = labels_2025,
    aes(label = label),
    hjust = -0.2, size = 3.2, fontface = "bold",
    show.legend = FALSE
  ) +

  facet_wrap(~Income_Group, ncol = 3) +

  scale_color_manual(values = col_measure, name = NULL) +
  scale_linetype_manual(values = c("Real Income" = "solid",
                                   "HDB Resale Price" = "solid"),
                        name = NULL) +
  scale_x_continuous(breaks = seq(2000, 2025, by = 5),
                     expand = expansion(mult = c(0.02, 0.14))) +
  scale_y_continuous(
    breaks = seq(60, 180, by = 20),
    limits = c(55, 185),
    sec.axis = sec_axis(
      ~ .,
      breaks = 100,
      labels = "← base year (2010 = 100)"
    )
  ) +

  labs(
    title    = "Real Household Income vs HDB Resale Price Index (2010 = 100)",
    subtitle = paste0(
      "Singapore, 2000–2025  |  Income deflated by group-specific CPI  |  ",
      "Shaded band = affordability gap  |  Labels show 2025 index values"
    ),
    x       = NULL,
    y       = "Index value (2010 = 100)",
    caption = paste0(
      "Sources: Department of Statistics Singapore — Average Monthly Household Employment Income by Decile; ",
      "CPI by Household Income Group (2024 base year).\n",
      "Housing & Development Board / data.gov.sg — HDB Resale Flat Prices (transaction records, 2000–2025).\n",
      "Note: Income groups deflated using their respective group-specific CPI. ",
      "HDB index computed as annual median resale price, rebased to 2010 = 100."
    )
  ) +

  base_theme +
  theme(
    legend.text      = element_text(size = 11),
    legend.spacing.x = unit(0.4, "cm"),
    # Style the secondary axis label
    axis.text.y.right = element_text(size = 8, color = "grey50",
                                     face = "italic", hjust = 0),
    axis.ticks.y.right = element_line(color = "grey55")
  )

p
```

**Reading the chart.** Each panel shows one income group. The blue line tracks real (inflation-adjusted) household income; the red line tracks HDB resale prices. Both are indexed to 2010 = 100, so the lines share a common baseline — any divergence directly reflects a change in relative affordability. The shaded band between the lines is the affordability gap: when red rises above blue, housing is outpacing income.

---

## Key Insights

**Before 2010**, real income broadly kept pace with or exceeded HDB resale prices across all three income groups, reflecting relative affordability stability in the early 2000s.

**From 2010 onward**, resale prices diverge upward. The gap widens progressively, then accelerates sharply after 2020 — coinciding with the post-COVID resale surge driven by demand brought forward from the pandemic period.

**By 2025**, the HDB index reaches 171.6 — its highest level in the 25-year window — while real income indices are 140.5 (Bottom 20%), 141.8 (Middle 60%), and 113.8 (Top 20%). The affordability gap is largest in absolute index-point terms for the Top 20% (57.8 points), but this reflects real income plateauing under CPI deflation rather than greater financial stress. **The Bottom 20% gap (31.1 points) is the most policy-relevant**: their lower absolute income provides far less buffer against rising prices, and their real income growth in 2020–2025 is the weakest of all three groups.

---

## Supporting Analysis

::: {.callout-note collapse="true"}
### Affordability gap over time
```{r gap-chart, fig.width=10, fig.height=5.5}
gap_labels_2025 <- gap_data %>%
  filter(Year == 2025) %>%
  mutate(
    label = paste0(Income_Group, ": +", round(Affordability_Gap, 1), " pts"),
    # Stagger y positions to avoid overlapping
    y_pos = case_when(
      Income_Group == "Top 20%"    ~ 59,
      Income_Group == "Bottom 20%" ~ 32,
      Income_Group == "Middle 60%" ~ 26
    )
  )

ggplot(gap_data, aes(x = Year, y = Affordability_Gap, color = Income_Group)) +
  geom_rect(data = events,
            aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
            inherit.aes = FALSE, fill = "grey50", alpha = 0.10) +
  # Event labels at bottom, rotated — same as main chart
  geom_text(data = events,
          aes(x = (xmin + xmax) / 2, y = -39, label = label),
          inherit.aes = FALSE,
          size = 2.5, color = "grey35", fontface = "italic",
          angle = 90) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.5) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  # Staggered endpoint labels
  geom_text(
    data = gap_labels_2025,
    aes(y = y_pos, label = label),
    x = 2026, hjust = 0, size = 3, fontface = "bold",
    show.legend = FALSE
  ) +
  scale_color_manual(values = col_group, name = NULL) +
  scale_x_continuous(
    breaks = seq(2000, 2025, by = 5),
    limits = c(1999, 2033),
    expand = expansion(mult = c(0, 0))
  ) +
  scale_y_continuous(
    breaks = seq(-40, 80, by = 20),
    limits = c(-42, 85)
  ) +
  labs(title    = "Affordability gap by income group",
       subtitle = "Gap = HDB index − Real income index  |  Positive = housing outpacing income",
       x = NULL, y = "Gap (index points)",
       caption  = "Sources: SingStat, HDB / data.gov.sg") +
  base_theme
```
:::

::: {.callout-note collapse="true"}
### CAGR by sub-period
```{r cagr-chart, fig.width=11, fig.height=5}
cagr_fn <- function(end_val, start_val, n) round(((end_val/start_val)^(1/n) - 1)*100, 2)

periods <- list(c(2000,2010), c(2010,2020), c(2020,2025))

income_cagr <- map_dfr(periods, function(p) {
  tidy_data %>%
    filter(Measure_Type == "Real Income", Year %in% p) %>%
    pivot_wider(names_from = Year, values_from = Index_Value) %>%
    mutate(CAGR = cagr_fn(.[[4]], .[[3]], p[2]-p[1]),
           Sub_Period = paste0(p[1],"–",p[2]),
           Measure = "Real Income") %>%
    select(Income_Group, Sub_Period, Measure, CAGR)
})

hdb_cagr <- map_dfr(periods, function(p) {
  vals <- tidy_data %>%
    filter(Measure_Type == "HDB Resale Price", Year %in% p) %>%
    distinct(Year, Index_Value) %>%
    pivot_wider(names_from = Year, values_from = Index_Value)
  tibble(CAGR = cagr_fn(vals[[2]], vals[[1]], p[2]-p[1]),
         Sub_Period = paste0(p[1],"–",p[2]),
         Measure = "HDB Resale Price")
}) %>% crossing(Income_Group = c("Bottom 20%","Middle 60%","Top 20%"))

cagr_data <- bind_rows(income_cagr, hdb_cagr) %>%
  mutate(Income_Group = factor(Income_Group, levels = c("Bottom 20%","Middle 60%","Top 20%")),
         Sub_Period   = factor(Sub_Period, levels = c("2000–2010","2010–2020","2020–2025")),
         Measure      = factor(Measure, levels = c("Real Income","HDB Resale Price")))

ggplot(cagr_data, aes(x = Sub_Period, y = CAGR, fill = Measure)) +
  geom_col(position = position_dodge(0.65), width = 0.55) +
  geom_hline(yintercept = 0, linewidth = 0.4, color = "grey40") +
  geom_text(aes(label = paste0(round(CAGR,1),"%"),
                vjust = ifelse(CAGR >= 0, -0.4, 1.3)),
            position = position_dodge(0.65), size = 2.8, fontface = "bold") +
  facet_wrap(~Income_Group, ncol = 3) +
  scale_fill_manual(values = col_measure, name = NULL) +
  scale_y_continuous(labels = function(x) paste0(x,"%"),
                     breaks = seq(-2, 10, by = 2)) +
  labs(title    = "CAGR of real income vs HDB resale prices by sub-period",
       subtitle = "Three periods: 2000–2010, 2010–2020, 2020–2025",
       x = NULL, y = "CAGR (%)",
       caption  = "Sources: SingStat, HDB / data.gov.sg") +
  base_theme + theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
```
:::