---
title: "Visualisation – SG Housing Affordability vs Income"
subtitle: "Did real income growth keep pace with housing price growth, 2000–2025?"
author:
  - name: Shina Shih Xin Rong (2402781)
  - name: Jeanie Cherie Chua Yue-Ning (2403083)
  - name: Claudia Yue Xin Ying (2401817)
  - name: Nurul Zahirah Binte Muhamadnoh (2401671)
  - name: Anushka Chourasia (2503769)
date: "2026-03-05"
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 2
    number-sections: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup

library(tidyverse)
library(knitr)
library(scales)

# Load the clean dataset produced by data-pipeline.qmd
tidy_data <- readRDS("tidy_data.rds")

# Derive gap dataset used by Charts 2 and 3
gap_data <- tidy_data %>%
  pivot_wider(names_from = Measure_Type, values_from = Index_Value) %>%
  mutate(Affordability_Gap = `HDB Resale Price` - `Real Income`)

# Shared colour palettes
col_measure  <- c("Real Income" = "#2166AC", "HDB Resale Price" = "#B2182B")
col_group    <- c("Bottom 20%" = "#D73027", "Middle 60%" = "#FC8D59", "Top 20%" = "#4575B4")

# Shared annotation helpers
gfc_rect   <- annotate("rect", xmin = 2008, xmax = 2009,
                        ymin = -Inf, ymax = Inf, alpha = 0.08, fill = "grey30")
covid_rect <- annotate("rect", xmin = 2020, xmax = 2021,
                        ymin = -Inf, ymax = Inf, alpha = 0.08, fill = "grey30")

shared_theme <- theme_minimal(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10, color = "grey40"),
    plot.caption  = element_text(size = 8,  color = "grey50"),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  )
```

## Chart 1 – Faceted indexed comparison {#sec-chart1}

*Real income vs HDB resale price growth, by income group*

```{r}
#| label: fig-faceted-indexed
#| fig-cap: "Real household income vs HDB resale price index (2010 = 100). Each panel shows one income group; the HDB resale price line (red) is identical across panels, making cross-group divergence directly comparable."
#| fig-width: 12
#| fig-height: 5

# Compute 2025 endpoint labels for annotation
labels_2025 <- tidy_data %>%
  filter(Year == 2025) %>%
  mutate(label = paste0(round(Index_Value, 0)))

ggplot(tidy_data, aes(x = Year, y = Index_Value, color = Measure_Type)) +
  # Shaded ribbon between the two lines (gap emphasis)
  geom_ribbon(
    data = gap_data,
    aes(x = Year, ymin = `Real Income`, ymax = `HDB Resale Price`),
    inherit.aes = FALSE,
    fill = "#B2182B", alpha = 0.08
  ) +
  gfc_rect +
  covid_rect +
  annotate("text", x = 2008.5, y = 145, label = "GFC",      size = 2.5, color = "grey40", angle = 90) +
  annotate("text", x = 2020.5, y = 145, label = "COVID-19", size = 2.5, color = "grey40", angle = 90) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey60", linewidth = 0.4) +
  geom_line(linewidth = 1) +
  # 2025 endpoint labels
  geom_text(
    data = labels_2025,
    aes(label = label),
    hjust = -0.15, size = 3, fontface = "bold",
    show.legend = FALSE
  ) +
  facet_wrap(~Income_Group, ncol = 3) +
  scale_color_manual(values = col_measure, name = NULL) +
  scale_x_continuous(breaks = seq(2000, 2025, by = 5), expand = expansion(mult = c(0, 0.12))) +
  scale_y_continuous(breaks = seq(60, 180, by = 20)) +
  labs(
    title    = "Real household income vs HDB resale price index (2010 = 100)",
    subtitle = "Singapore, 2000–2025  |  Income deflated by group-specific CPI  |  Shaded band = affordability gap",
    x = NULL, y = "Index (2010 = 100)",
    caption  = "Sources: SingStat (household income, CPI by income group), HDB (resale transaction prices)"
  ) +
  shared_theme +
  theme(
    strip.text   = element_text(face = "bold", size = 11),
    axis.text.x  = element_text(angle = 45, hjust = 1, size = 8)
  )
```

Prior to 2010, real income across all three groups broadly tracked or outpaced HDB resale price growth, indicating relative affordability stability. From 2010 onward, resale prices diverge upward. The gap accelerates sharply after 2020, coinciding with the post-COVID resale surge. By 2025, the HDB index reaches its highest recorded level while real income growth plateaus for all groups — most visibly for the Bottom 20%.

---

## Chart 2 – Affordability gap over time {#sec-chart2}

*How far housing price growth has outpaced real income growth, per group*

```{r}
#| label: fig-affordability-gap
#| fig-cap: "Affordability gap = HDB resale price index minus real income index (2010 = 100). Positive values indicate housing prices growing faster than income. Annotated 2025 gap values are shown at the right margin."
#| fig-width: 10
#| fig-height: 5.5

# 2025 endpoint labels for gap chart
gap_labels_2025 <- gap_data %>%
  filter(Year == 2025) %>%
  mutate(label = paste0(Income_Group, ": ", round(Affordability_Gap, 1), " pts"))

ggplot(gap_data, aes(x = Year, y = Affordability_Gap, color = Income_Group)) +
  gfc_rect +
  covid_rect +
  annotate("text", x = 2008.5, y = 68, label = "GFC",      size = 2.5, color = "grey40", angle = 90) +
  annotate("text", x = 2020.5, y = 68, label = "COVID-19", size = 2.5, color = "grey40", angle = 90) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.5) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.5) +
  # Endpoint labels
  geom_text(
    data = gap_labels_2025,
    aes(label = label),
    hjust = -0.05, size = 3, fontface = "bold",
    show.legend = FALSE
  ) +
  scale_color_manual(values = col_group, name = NULL) +
  scale_x_continuous(breaks = seq(2000, 2025, by = 5), expand = expansion(mult = c(0, 0.25))) +
  scale_y_continuous(breaks = seq(-20, 80, by = 20)) +
  labs(
    title    = "Housing affordability gap by income group (2010 = 100)",
    subtitle = "Gap = HDB resale price index − real income index  |  Positive = housing outpacing income",
    x = NULL, y = "Affordability gap (index points)",
    caption  = "Sources: SingStat (household income, CPI by income group), HDB (resale transaction prices)"
  ) +
  shared_theme
```

Before 2010, all three groups show near-zero or negative gaps — income was keeping pace with prices. From 2010, gaps turn persistently positive. The steepest widening occurs post-2020. By 2025, the Top 20% shows the largest index-point gap (driven by real income plateauing under CPI deflation), but this does not imply greater financial stress — higher absolute incomes and better financing access buffer that group. The Bottom 20% gap is the most policy-relevant, as lower absolute incomes leave less buffer against rising prices.

---

## Chart 3 – CAGR comparison across sub-periods {#sec-chart3}

*Where did growth occur, and which sub-period was most damaging for affordability?*

```{r}
#| label: fig-cagr
#| fig-cap: "Compound annual growth rate (CAGR) of real income and HDB resale prices across three sub-periods. Bars above the dashed zero line indicate positive growth; the gap between the HDB bar and the income bar in each period reflects the direction of affordability change."
#| fig-width: 11
#| fig-height: 5.5

# Helper: CAGR between two years
cagr <- function(end_val, start_val, n_years) {
  round((end_val / start_val)^(1 / n_years) - 1, 4) * 100
}

# Income CAGRs per group and sub-period
periods <- list(
  c(2000, 2010),
  c(2010, 2020),
  c(2020, 2025)
)

income_cagr <- map_dfr(periods, function(p) {
  vals <- tidy_data %>%
    filter(Measure_Type == "Real Income", Year %in% p) %>%
    pivot_wider(names_from = Year, values_from = Index_Value)

  vals %>%
    mutate(
      CAGR        = cagr(.[[4]], .[[3]], p[2] - p[1]),
      Sub_Period  = paste0(p[1], "–", p[2]),
      Measure     = "Real Income"
    ) %>%
    select(Income_Group, Sub_Period, Measure, CAGR)
})

# HDB CAGR per sub-period (same for all groups — replicate for faceting)
hdb_cagr <- map_dfr(periods, function(p) {
  vals <- tidy_data %>%
    filter(Measure_Type == "HDB Resale Price", Year %in% p) %>%
    distinct(Year, Index_Value) %>%
    pivot_wider(names_from = Year, values_from = Index_Value)

  tibble(
    CAGR       = cagr(vals[[2]], vals[[1]], p[2] - p[1]),
    Sub_Period = paste0(p[1], "–", p[2]),
    Measure    = "HDB Resale Price"
  )
}) %>%
  crossing(Income_Group = c("Bottom 20%", "Middle 60%", "Top 20%"))

cagr_data <- bind_rows(income_cagr, hdb_cagr) %>%
  mutate(
    Income_Group = factor(Income_Group, levels = c("Bottom 20%", "Middle 60%", "Top 20%")),
    Sub_Period   = factor(Sub_Period,   levels = c("2000–2010", "2010–2020", "2020–2025")),
    Measure      = factor(Measure,      levels = c("Real Income", "HDB Resale Price"))
  )

ggplot(cagr_data, aes(x = Sub_Period, y = CAGR, fill = Measure)) +
  geom_col(position = position_dodge(width = 0.65), width = 0.55) +
  geom_hline(yintercept = 0, linewidth = 0.4, color = "grey40") +
  geom_text(
    aes(label = paste0(round(CAGR, 1), "%"),
        vjust = ifelse(CAGR >= 0, -0.4, 1.3)),
    position = position_dodge(width = 0.65),
    size = 2.8, fontface = "bold"
  ) +
  facet_wrap(~Income_Group, ncol = 3) +
  scale_fill_manual(values = col_measure, name = NULL) +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     breaks = seq(-2, 10, by = 2)) +
  labs(
    title    = "CAGR of real income vs HDB resale prices by sub-period (2010 = 100 base)",
    subtitle = "Singapore  |  Three periods: pre-GFC baseline, post-GFC expansion, COVID and post-pandemic surge",
    x = NULL, y = "CAGR (%)",
    caption  = "Sources: SingStat (household income, CPI by income group), HDB (resale transaction prices)"
  ) +
  shared_theme +
  theme(
    strip.text  = element_text(face = "bold", size = 11),
    axis.text.x = element_text(size = 9)
  )
```

The 2000–2010 period shows income and housing growing at broadly comparable rates across all groups. The 2010–2020 period reveals a first divergence — HDB prices accelerate while income growth slows, particularly for the Bottom 20%. The 2020–2025 period is the most damaging: HDB resale price growth surges to its highest CAGR across all three sub-periods while real income growth falls sharply — in the Bottom 20%, real income CAGR turns near-zero or negative, directly explaining the widening gap seen in Charts 1 and 2.

---

## Chart 4 – Affordability gap summary table {#sec-chart4}

*Snapshot of gap values at key years, per income group*

```{r}
#| label: tbl-gap-summary
#| tbl-cap: "Affordability gap (HDB resale price index − real income index, 2010 = 100) at selected years. Positive values indicate housing prices outpacing real income growth."

key_years <- c(2000, 2005, 2010, 2015, 2020, 2023, 2025)

gap_table <- gap_data %>%
  filter(Year %in% key_years) %>%
  select(Year, Income_Group, `HDB Resale Price`, `Real Income`, Affordability_Gap) %>%
  mutate(across(c(`HDB Resale Price`, `Real Income`, Affordability_Gap), ~ round(., 1))) %>%
  pivot_wider(
    names_from  = Income_Group,
    values_from = Affordability_Gap,
    id_cols     = c(Year, `HDB Resale Price`)
  ) %>%
  rename(
    "HDB index" = `HDB Resale Price`,
    "Gap: Bottom 20%" = `Bottom 20%`,
    "Gap: Middle 60%" = `Middle 60%`,
    "Gap: Top 20%"    = `Top 20%`
  )

kable(
  gap_table,
  align = c("l", rep("r", ncol(gap_table) - 1)),
  caption = NULL
)
```

The table confirms that all groups had near-zero or negative gaps before 2010, meaning income growth was outpacing housing prices. Gaps turn positive around 2011–2012 and widen continuously through 2025. The 2020–2025 acceleration is visible across all three columns, with the Bottom 20% gap narrower in absolute points than the Top 20% but more consequential given their lower absolute income base.

---

## Key findings {#sec-findings}

Taken together, the four charts support the following answers to the research question:

**Did real income growth keep pace with housing prices?** No, not after 2010. Prior to 2010, real income growth broadly matched or exceeded HDB resale price growth across all income groups. Post-2010 — and especially post-2020 — housing prices have consistently outpaced real income growth.

**Have affordability gaps widened across income groups?** Yes. All three groups face larger affordability gaps in 2025 than at any prior point in the 2000–2025 window. The rate of widening accelerated sharply after 2020. The Bottom 20% gap is the most policy-relevant: their lower absolute income provides less buffer against price growth, and their real income CAGR in the 2020–2025 period is the weakest of all three groups.

**Limitations to note when reading these charts.** Group-specific CPI deflation (used here) is more accurate than a single all-items deflator, but still may understate true housing cost burden for the Bottom 20%. The HDB index reflects transaction prices only and does not capture mortgage servicing costs, which rose sharply in 2022–2024 as interest rates increased. Gross income including employer CPF contributions is used throughout, which overstates disposable income available for housing payments.
